# R Basics

::: callout-caution
### Under Construction
:::

What follows assumes you have installed either RStudio or Positron, though almost everything discussed also applies to the base R environment that comes with the language on Windows and Macintosh computers.

## The Console Pane

When you launch the base-R environment (for Windows or Mac), launch RStudio, or launch Positron, somewhere on the screen will be a window pane labelled "*Console*". Assuming you have not tweaked the default layout, all three environments will have this located on the left side (see the sections highlighted in red in @fig-IDE_panes_html).

::: {.content-visible when-format="html" #fig-IDE_panes_html layout-nrow=2}

![RStudio](imgs/basics/rstudio_default_panes.svg){#fig-rstudio_panes_html}

![Positron](imgs/basics/positron_default_panes.svg){#fig-positron_panes_html}

RStudio (a) and Positron (b) integrated development environments. Each pane serves a specific purpose for writing, running, and managing R code. Note that the scripting pane (top left) does not appear by default—open it by selecting File → New File.
:::

::: {.content-visible when-format="pdf" layout-ncol=2}

![RStudio](imgs/basics/rstudio_default_panes.pdf){#fig-rstudio_panes_html}

![Positron](imgs/basics/positron_default_panes.pdf){#fig-positron_panes_pdf}

Test
:::

The console pane functions as a terminal for the R language specifically.[^console_py] Though, to distinguish it from your operating system's main terminal it is labelled "console." 

Within the console pane, you should see a $>$. This symbol denotes the *command line's prompt*.  In other words, it denotes the space in which you type commands, using R *code*, to your computer (this is the section highlighted in purple in @fig-IDE_panes_html). The term "code" here is just a shorthand way of referring to "computer code" which is a more modern way of expressing the fact that we are typing commands using a programming language. The presence of $>$ indicates that the computer is awaiting your command.

[^console_py]:This is technically a lie. RStudio and Positron give you the ability to toggle between both R and Python.

If you type `1 + 1` on the command line and then press "*enter/return*" on your keyboard, you should see a `2` display as an *output* almost instantaneously beneath it. In this case the expression `1 + 1` is a line of R code.  Pressing *enter/return*, runs (or executes) this R code. The `2` is the computer's resulting *output*."

Input:
```r
1 + 1
```

Output:
```{r}
#| echo: false
1 + 1
```

If you close RStudio or Positron, you will find that any history of this calculation is gone when you re-open the environment.  Consequently, typing commands into the console offers us a quick way to perform simple tasks that we are not necessarily concerned with preserving.  However, in most cases we will be typing R code that we do want to preserve, run, edit, and add to at later date.  This is where the concept of a *script* becomes important.

## The Scripting Pane

A script is a text document where you can type, run, edit, and save your R code. To create a new script in RStudio, select the *File* menu in the top left corner, hover over *New File*, and choose *R Script*. In Positron, the process is identical except you select *R File* instead.

Once opened, you can type R code into this new pane and save it in the conventional manner of most word processing applications (i.e., *File* $\rightarrow$ *Save As*). Additionally, this pane permits you to run lines of code selectively or all together. For instance, if you type the following into the script pane ...

```r
1 + 1
2 + 2
3 + 3
```

You can now place your cursor at a line of your choosing and run that line individually.  To do this in RStudio or Positron simply hold down the Control key on your keyboard and press *Enter/Return*. If you highlight all the lines of code, or just a subset of them, you can then run that highlighted section in a similar manner.

## Keyboard Shortcuts {#sec-key_shorts}
It is at this juncture that a noteworthy feature of programming environments be mentioned; specifically, keyboard shortcuts (also called "hotkeys"). All robust programming environments equip users with the ability to perform virtually any non-typing task directly from the keyboard, increasing efficiency and comfort. 

For instance, if you are using the Windows operating system, holding down the "*control*" key and pressing the "*s*" key will save your script file (Ctrl + S). Learning the shortcuts for frequently used features, such as selecting and running lines of code, will make the process of writing code considerably more time efficient and effortless.  In theory, a good programmer - using a competently developed coding environment - should never require the use of a mouse.  

RStudio and Positron both offer a wide range of keyboard shortcuts that can be customized to user preferences.  In RStudio, selecting *Help* $\rightarrow$ *Keyboard Shortcuts Help* will display a list of existing shortcuts that users can avail themselves of.  The same can be done in Positron by selecting *File* $\rightarrow$ *Preferences* $\rightarrow$ *Keyboard Shortcuts*. 

Please note, it is not being suggested that you go out of your way to memorize all of these at once. The simple act of trying to use them consistently will be sufficient to learn them in an effortless manner. At the outset, it is to your advantage to merely select a few and attempt to use them consistently while you code. A few of the most useful ones are listed in @tbl-key_shorts.  Many of these are not even exclusive to RStudio or Positron, but are just standard shortcuts in most operating systems.[^key_shorts]

| Description | Windows/Linux | Macintosh |
|-------------|---------|-----------|
| Run current line/section | Ctrl + Enter | Cmd + Return |
| Clear Console | Ctrl + L | Ctrl + L |
| Move to the beginning of a line | Home | Cmd + Left |
| Move to the end of a line | End | Cmd + Right |
| Move the cursor one word/block at a time | Ctrl + Left or Right | Option + Left or Right |
| Highlight all | Ctrl + A | Cmd + A |
| Highlight sections | Shift + Up, Down, Left, or Right | Shift + Up, Down, Left, or Right |
| Move cursor to script window | Ctrl + 1 | Ctrl + 1 |
| Move cursor to console window | Ctrl + 2 | Ctrl + 2 |
| Type the `<-` operator | Alt + - (minus) | Option + - (minus) |

: Useful Keyboard Shortcuts {#tbl-key_shorts}

[^key_shorts]:Shortcuts 3, 4, and 5 can be combined with shortcut 7 to highlight bigger sections of code.

::: callout-important
When utilizing the keyboard shortcuts mentioned in @sec-key_shorts, it is worth remembering that standard QWERTY-style keyboards are symmetrically designed. Modifier keys like the *shift* key, *control* key, and *alt* key are located on both the left and right side of the board.[^key_layout] This is not by accident and many people - even those who have grown up with unprecedented access to computers and the internet - have never learned to appreciate the utility of this layout or use it appropriately.

As an example, to type capital letters you should always depress the shift key on the opposite side of the keyboard to the letter. So, if you desired to type the capital letter *Q*, you would depress the right shift key with your right hand, and type *Q* with your left hand.  A similar logic applies to the other modifier keys. As another example, to use the keyboard shortcut in row 9 of @tbl-key_shorts, you would depress the right control key (with your right hand) and use your left hand to press the *2* key. You should not be trying to press both keys with a single hand.  Such advice might seem obvious but, given the sheer number of people who contort their wrists and fingers in grotesquely strange and painful ways, it is clearly far from being so.
:::

[^key_layout]:If this isn't true of your keyboard, it's time to get a better keyboard.

## Basic Arithmetic

At its core R is really nothing more than a powerful calculator, and we can use it as such.  R can be used to add ($+$), subtract ($-$), multiply ($\times$), and divide ($\div$).

```{r results = "hold"}
666 + 13
13 - 666
9 * 27
666 / 9
```

Exponents can be incorporated as well by using the $^\wedge$ ("caret"), symbol. For instance, the expression $9^3$ can be written as ...

```{r}
9^3
```

R will also follow the ritualistic *order of operations* when dealing with more complex expressions.  To illustrate, consider the mathematical statement $8\div2(2+2)$. Some people mistakenly believe that this expression is equal to $1$, some believe it is equal to $4$, and others believe that it is improperly written and there is no solution. In fact, it is equal to $16$. As many will no doubt have learned in their primary education, according to order of operations (BEDMAS[^BEDMAS]), the order in which you divide and multiply inside the equation is not fixed, sometimes you divide first and sometimes you multiply first. However, what most people never learn is that the order you use is not up to you. You must always calculate from left to right when making a choice between multiplication and division.  The same rule applies to addition and subtraction.

[^BEDMAS]:"BEDMAS" of course being the famous mnemonic to help memorize the order of operations: Brackets, Exponents, Division, Multiplication, Addition, and Subtraction. Many non-Canadian readers may be more familiar with the inferior variants of this mnemonic, PEDMAS and PEMDAS.

```{r}
8/2*(2+2)
```

If we re-write the equation to be $8\div(2+2)2$, you will see a corresponding change in the computer's output.

```{r}
8/(2+2)*2
```

R also has the ability to perform *Euclidean Division,* which many may recall from their long suffering days in primary education days as simply "division with a remainder."  For instance, consider $11 \div 2$.  Conventionally, you would want and expect an answer of $5.5$, and R will produce that.

```{r}
11/2
```

However, if we want to see the result expressed as a quotient and remainder (i.e., if we want to use Euclidean Division), we could obtain the quotient by typing ...

```{r}
11 %/% 2
```

To obtain the remainder we type...
```{r}
11 %% 2
```

Thus, $11$ can be split into $2$ groups of $5$, with $1$ left over.  More technically, the `%%` is what is known as the *modulo operator* and the remainder value of `1` that results from `11 %% 2` is known as the *modulus*.

Other, more complex, arithmetic operations are available in the R language; however, most of them will require the use of specialized lines of code called *functions*, which are discussed later.

Given that we are on the topic of basic arithmetic, it is perhaps worth considering what happens when you "break the rules" of basic arithmetic. Suppose we divide a positive and negative value by zero, what will happen?

```{r}
1/0
-1/0
```

You can see that R produces a result of `Inf` and `-Inf` which is an abbreviated way of referring to infinity ($\infty$) in the positive and negative directions respectively.[^large_num]

[^large_num]:This will also be generated if a number is too large for a computer to cope with. For example, the code `.Machine$double.xmax` will produce the largest number your computer can handle. R will technically still let you add values to this number, but the number won't appear to change because the amount you would have to add to alter what is shown is excessively large. However, if you multiply it by $2$, you should get `Inf`.

What happens if you take the square root of a negative number?[^sqrt_neg]

[^sqrt_neg]:Recall that exponents can be used to take the square-root of a number.  For example, $\sqrt{4}$ can be expressed as $4^\frac{1}{2}$.

```{r}
(-4)^(1/2)
```

The abbreviation `NaN` here stands for `not a number,` and is a fairly sensible output given that the square root of a negative number does not exist as a real number (it only exists in your imagination).

Finally, since its use crops up from time to time, it can be handy to know that R comes with the number $\pi$ stored as a constant. To use it, you need only type `pi`.[^mmmm_pi]

[^mmmm_pi]:If you find $\pi$ displayed to seven digits inadequate, you may want to talk to a professionally licensed therapist. Alternatively, you can display more digits by running the code `print(pi, digits = 16)`. Values exceeding $16$ digits will be inaccurate given the limitations of 64-bit computers, so it is advisable not to go beyond $16$ even though a max of $22$ are possible. If you want R to always display all $16$ digits, you can change its default behaviour by running `options(digits = 16)`, though this is not recommended.

```{r}
pi
```

## Scientific Notation

On occasion values will be either excessively large or excessively small. In such cases R will often display the values using what is referred to as *scientific notation*.  For instance, dividing the number $2$ by $100000$ will result in scientific notation being employed:

```{r}
2 / 100000
```

Notice the `e-05` in the output. This is how you know R is presenting a number using scientific notation. To interpret this in a conventional manner, imagine there is a decimal point after the $2$, like so: `2.0e-05`. Then just move that decimal point five digits to the left. In other words, `2e-05` is the same as writing `0.00002`. Mathematically, `2e-05` translates to $2 \times 10^{-5}$.

If the output were showing `e+5`, then you would move the decimal five digits to the right. For example, `2e+5` is the same as writing `200000`. Notice there are five 0s; this is because, mathematically, `2e+5` means $2 \times 10^5$

Remember that positive powers move the decimal right (in the positive direction), and negative powers move the decimal left (in the negative direction).

## Commenting Out Lines

In the course of writing R code, there will be occasions where you would like to run a script you have typed up, but not necessarily run every single line on that script.  There might be certain lines that you would, at least tentatively, like to keep for one reason or another but not necessarily run.  You can accomplish this by "commenting out" your code.  If you type a `#` symbol, any code that follows that symbol and is on the same line as that symbol will not be run.

```{r results = "hold"}
1 + 5
# 2 + 4
3 + 3
1 + 2 + 3 # + 4 + 5
```

This process is phrased "commenting out" because using the `#` is also frequently employed to write short helpful comments to yourself and other readers about your R script.

## Creating Objects

A central feature of R is its ability to call objects in memory. For instance, we can define an object name, `x`, and have that name represent a number by typing a little arrow, `<-`, and following it with a value such as $1$.

```{r}
x <- 1
```

You will find that running this line of code produces no corresponding output.  However, if we now run `x` by itself the computer will display an output of $1$.

```{r}
x
```

If you look into how R actually stores what we have done in memory, the "object" in memory is the number $1$.  `x` is merely a name we are assigning to that object. However, a lot of R users are under the impression that the reverse is true - i.e., that we have in some sense created an object called `x` and stored something inside of it, but that is not actually the case. `x` is just a name binded to the object $1$, and this object $1$ is located somewhere inside your computer's memory. Admittedly, unless you are doing some seriously advanced R programming, this is a distinction that will not matter to most R users, but it is important because it means that if you do something like this . . . 

```{r}
x <- 1
y <- 1
```

`x` and `y` are technically different objects in the computer's memory.  However, if we did this ....

```{r}
y <- x
```

they now represent the same object in memory. Moreover, altering one does not affect the other and just ends up creating two separate objects in memory. E.g. ...

```{r results = "hold"}
x <- x + 1
x
y
```

To see a complete list of objects presently loaded in memory have a look at the *Environment* window pane in R studio and the *Variables* pane in Positron.