---
execute:
  fig-width: 6
  fig-height: 5
---

# Plotting Basics with the Tidyverse {#sec-plotting}

The history of R can be split into two epochs. First, there was the age before the *tidyverse*. This was a time of considerable friction, where data analysts wrestled with inconsistent syntax and workflows that seemed to resist them at every turn. It was a formative era, and the frustrations it produced planted the seeds for everything that would follow.

::: {style="float: right; margin: 0 0 1em 1em; max-width: 35%;"}
![An engraving depicting acolytes of the tidyverse burning live sacrifices, captive within a large wicker effigy, to appease their deities [@Pennant1784].](imgs/plotting/t_pennant.png){#fig-wicker fig-alt="tidyverse's wickerman"}
:::

Then came the *tidyverse*. A genuine turning point in the R world. As described by its website (<a href = "https://www.tidyverse.org/" target = "_blank">https://www.tidyverse.org/</a>), the *tidyverse* is an opinionated collection of R packages (see @sec-R_packages), brought to life by Hadley Wickham and a dedicated community of contributors [@Wickham2019]. Where once working with data in R demanded patience and hard-won expertise, the tidyverse made everything much more accessible and efficient. This meant that anyone willing to learn, could develop the skills to wrangle, transform, and visualise data with clarity and confidence. However, not everyone welcomed it with open arms, and some in the R community viewed the new approach with scepticism [@Muenchen], but the tidyverse has since become a cornerstone of modern R practice and transformed an already elegant language into something genuinely magical. 

Each package within the tidyverse can be installed individually, though most people find it simplest to install them all at once with a single command, which is exactly where we will begin. From there, this chapter will introduce you to *ggplot2*, the tidyverse's powerful and elegant system for visualizing data. By the end, you will be producing real plots from real data, and you will have a framework for how *ggplot2* functions that will serve you well throughout the rest of this book.

## Installing the tidyverse {#sec-tidy_install}

Each package within the tidyverse can be installed individually, though most find it easiest to install every package within the scope of the tidyverse all at once.

```r
install.packages("tidyverse")
```

```{r}
# Load the packages
library(tidyverse)
```

While the above code installs all the packages, running `library(tidyverse)` only loads the the nine "core" packages:  *dplyr*, *readr*, *forcats*, *stringr*, **ggplot2**, *tibble*, *lubridate*, *tidyr*, and *purr*. Other tidyverse packages, such as *readxl*, will need to be loaded separately using the `library()` function.

::: {.callout-note collapse="true" icon = false}
## <i class="bi bi-tux"></i> A Note for Linux Users
As mentioned in @sec-pkg_depend, Linux tends to follow a philosophy of minimal, modular installations. That means system libraries are not bundled with the base OS, and installing the *tidyverse* will likely fail unless the libraries it depends on are already present on your system.

Before running `install.packages("tidyverse")`, open a terminal and run the appropriate commands for your distribution to install the required system packages. 

**Debian, Ubuntu, and Ubuntu-based distributions (Kubuntu, Mint, etc.):**

```bash
sudo apt update
sudo apt install -y \
  libcurl4-openssl-dev \
  libssl-dev \
  libxml2-dev \
  libfontconfig1-dev \
  libfreetype6-dev \
  libharfbuzz-dev \
  libfribidi-dev \
  libcairo2-dev \
  libpng-dev \
  libjpeg-dev \
  libtiff5-dev \
  libwebp-dev \
  zlib1g-dev
```

At the time of writing, this list should be sufficient for a successful *tidyverse* installation on Debian/Ubuntu-based systems. However, dependencies can change over time, so do not panic if something is missing. R's console output is usually quite informative about what is needed if the *tidyverse* install fails. In most cases it is as simple as running `sudo apt install <package-name>` (or the equivalent for your distribution). 

Users on other distributions (Fedora, Arch, openSUSE, etc.) will need the equivalent packages for their package manager. If the installation fails, R's console output will usually identify exactly what's missing.
:::

Speaking for the beginner, it will be noticed that when the tidyverse is loaded, not only is there a confirmation of what packages (and their versions) have been loaded, but there is also a list of "conflicts" displayed in the output.[^tidy_info]

[^tidy_info]:Most packages will not display this information for you quite so nicely as the tidyverse does, so pay attention to any messages you receive using the `library()` function.

For instance, two functions from the *dplyr* package, `filter()` and `lag()`, have the same name as pre-existing functions within R and, when you load a package with a conflict like this, precedence is always given to the most recently loaded package. This means, when you use the `filter()` function for example, R is going to use the version belonging to *dplyr*, not the original version that was a part of base R's *stats* package (which is pre-loaded each time you use R). Though you can still use that original version in the following manner: `package_name::function_name()`.  For example, `stats::filter()`. 

As a whole, the *tidyverse* will not solve all your problems, but it will come damn close. Admittedly, and this is particularly true for beginners, much of what the *tidyverse* offers will not be needed in your daily programming rituals, but will come in handy when least expected.

## Plotting with R

A core component of any GOOD DATA ANALYSIS obviously involves visualizing your data. As you progress through the various topics in this book, specific types of plots and their uses will be discussed in detail; however, for the time being, it will be helpful to get an intuitive sense of how plotting works with R generally. Thus, what follows in this section is intended to help you understand the logic of plotting with R. The goal at this point is not to make you an expert; rather, it is to provide beginners with a base level of knowledge and experience.

By itself, base R comes with a stock set of functions for plotting data. To illustrate we can run the following code to produce a nice looking histogram ...

```{r}
x <- rnorm(10000)
hist(x)
```

In the case of the above code, the function `rnorm()` is just generating $10,000$ random values.[^rnorm]

[^rnorm]:The random values are technically coming from a "standard normal" distribution (hence the "norm" in `rnorm`), but don't worry about that for now. The function `hist(x)`, is simply plotting those values as a histogram.  Running the code should generate an output similar to what you see below.

R's base plotting functions provide a convenient way to produce simple, high-quality plots, and they can be quite efficient when working with basic **univariate data** involving only one or two variables. However, modern research often demands the handling of far more complex datasets — ones that may include multiple **response variables** alongside numerous explanatory variables. Each additional variable adds layers of complexity and nuance to your data and, by extension, to the plots used to visualize them. While R's built-in plotting tools can accommodate these more elaborate scenarios, doing so often requires a high level of fluency with R's syntax and customization options. For this reason, this book will forgo R's base plotting system in favour of the widely respected *ggplot2* package, a core component of the *tidyverse*. *ggplot2* provides a more consistent and powerful framework for building complex plots, making it the preferred tool for data visualization throughout this text.

The "gg" in *ggplot2* stands for "grammar of graphics," a term borrowed from Leland Wilkinson's influential book of the same name [@Wilkinson2005]. Now, the word "grammar" might dredge up long-buried memories of dull English classes, but fear not. In this context, the term simply emphasizes that *ggplot2* is built on a coherent set of rules for assembling visualizations. This structure allows users to build a wide variety of plots in a consistent, modular way that can be easily tailored to their data and needs. This is a major improvement over many traditional plotting systems, which often require you to awkwardly cram your data into rigid, predefined formats—like trying to fit a square peg (your beautifully weird data) into a round hole (the software's narrow expectations).

The easiest way to understand how *ggplot2* works is to simply dive in and use it. Along the way, we will also learn a little bit more about R and data manipulation. However, a disclaimer is perhaps useful here:


::: callout-warning
This chapter contains a large variety of functions and strategies for plotting data with *ggplot2*. The reader would do well to head the advice provided at the start of @sec-warning.
:::

The first thing to do will be to ensure that *ggplot2* has been installed into our computer's library of packages and loaded so we can access its functions. As mentioned in @sec-tidy_install, if you have installed and loaded the *tidyverse*, this is already done, but if you chose not to do that,[^shame] *ggplot2* can be installed and loaded as a standalone package as well.

[^shame]:Shame on you.

<img src = "imgs/misc/under_construction.gif">