---
knitr:
  opts_chunk:
    fig.width: 3
    fig.height: 2.25
    out-width: "60%"
    fig.align: "center"
filters:
  - line-highlight
---

# Plotting Basics with the Tidyverse {#sec-plotting}

::: callout-caution
### Under Construction
:::

The history of R can be split into two epochs. First, there was the age before the *tidyverse*. This was a time of considerable friction, where data analysts wrestled with inconsistent syntax and workflows that seemed to resist them at every turn. It was a formative era, and the frustrations it produced planted the seeds for everything that would follow.

::: {style="float: right; margin: 0 0 1em 1em; max-width: 35%;"}
![An engraving depicting acolytes of the tidyverse burning live sacrifices, captive within a large wicker effigy, to appease their deities [@Pennant1784].](imgs/plotting/t_pennant.png){#fig-wicker fig-alt="tidyverse's wickerman"}
:::

Then came the *tidyverse*. A genuine turning point in the R world. As described by its website (<a href = "https://www.tidyverse.org/" target = "_blank">https://www.tidyverse.org/</a>), the *tidyverse* is an opinionated collection of R packages (see @sec-R_packages), brought to life by Hadley Wickham and a dedicated community of contributors [@Wickham2019]. Where once working with data in R demanded patience and hard-won expertise, the tidyverse made everything much more accessible and efficient. This meant that anyone willing to learn, could develop the skills to wrangle, transform, and visualise data with clarity and confidence. However, not everyone welcomed it with open arms, and some in the R community viewed the new approach with scepticism [@Muenchen], but the tidyverse has since become a cornerstone of modern R practice and transformed an already elegant language into something genuinely magical. 

Each package within the tidyverse can be installed individually, though most people find it simplest to install them all at once with a single command, which is exactly where we will begin. From there, this chapter will introduce you to *ggplot2*, the tidyverse's powerful and elegant system for visualizing data. By the end, you will be producing real plots from real data, and you will have a framework for how *ggplot2* functions that will serve you well throughout the rest of this book.

## Installing the tidyverse {#sec-tidy_install}

Each package within the tidyverse can be installed individually, though most find it easiest to install every package within the scope of the tidyverse all at once.

```r
install.packages("tidyverse")
```

```{r}
# Load the packages
library(tidyverse)
```

While the above code installs all the packages, running `library(tidyverse)` only loads the the nine "core" packages:  *dplyr*, *readr*, *forcats*, *stringr*, **ggplot2**, *tibble*, *lubridate*, *tidyr*, and *purr*. Other tidyverse packages, such as *readxl*, will need to be loaded separately using the `library()` function.

::: {.callout-note collapse="true" icon = false}
## <i class="bi bi-tux"></i> A Note for Linux Users
As mentioned in @sec-pkg_depend, Linux tends to follow a philosophy of minimal, modular installations. That means system libraries are not bundled with the base OS, and installing the *tidyverse* will likely fail unless the libraries it depends on are already present on your system.

Before running `install.packages("tidyverse")`, open a terminal and run the appropriate commands for your distribution to install the required system packages. 

**Debian, Ubuntu, and Ubuntu-based distributions (Kubuntu, Mint, etc.):**

```bash
sudo apt update
sudo apt install -y \
  libcurl4-openssl-dev \
  libssl-dev \
  libxml2-dev \
  libfontconfig1-dev \
  libfreetype6-dev \
  libharfbuzz-dev \
  libfribidi-dev \
  libcairo2-dev \
  libpng-dev \
  libjpeg-dev \
  libtiff5-dev \
  libwebp-dev \
  zlib1g-dev
```

At the time of writing, this list should be sufficient for a successful *tidyverse* installation on Debian/Ubuntu-based systems. However, dependencies can change over time, so do not panic if something is missing. R's console output is usually quite informative about what is needed if the *tidyverse* install fails. In most cases it is as simple as running `sudo apt install <package-name>` (or the equivalent for your distribution). 

Users on other distributions (Fedora, Arch, openSUSE, etc.) will need the equivalent packages for their package manager. If the installation fails, R's console output will usually identify exactly what's missing.
:::

Speaking for the beginner, it will be noticed that when the tidyverse is loaded, not only is there a confirmation of what packages (and their versions) have been loaded, but there is also a list of "conflicts" displayed in the output.[^tidy_info]

[^tidy_info]:Most packages will not display this information for you quite so nicely as the tidyverse does, so pay attention to any messages you receive using the `library()` function.

For instance, two functions from the *dplyr* package, `filter()` and `lag()`, have the same name as pre-existing functions within R and, when you load a package with a conflict like this, precedence is always given to the most recently loaded package. This means, when you use the `filter()` function for example, R is going to use the version belonging to *dplyr*, not the original version that was a part of base R's *stats* package (which is pre-loaded each time you use R). Though you can still use that original version in the following manner: `package_name::function_name()`.  For example, `stats::filter()`. 

As a whole, the *tidyverse* will not solve all your problems, but it will come damn close. Admittedly, and this is particularly true for beginners, much of what the *tidyverse* offers will not be needed in your daily programming rituals, but will come in handy when least expected.

## Plotting with R

A core component of any GOOD DATA ANALYSIS obviously involves visualizing your data. As you progress through the various topics in this book, specific types of plots and their uses will be discussed in detail; however, for the time being, it will be helpful to get an intuitive sense of how plotting works with R generally. Thus, what follows in this section is intended to help you understand the logic of plotting with R. The goal at this point is not to make you an expert; rather, it is to provide beginners with a base level of knowledge and experience.

By itself, base R comes with a stock set of functions for plotting data. To illustrate we can run the following code to produce a nice looking histogram ...

```{r}
#| fig-width: 5
#| fig-height: 4
#| out-width: "80%"
x <- rnorm(10000)
hist(x)
```

In the case of the above code, the function `rnorm()` is just generating $10,000$ random values.[^rnorm]

[^rnorm]:The random values are technically coming from a "standard normal" distribution (hence the "norm" in `rnorm`), but don't worry about that for now. The function `hist(x)`, is simply plotting those values as a histogram.  Running the code should generate an output similar to what you see below.

R's base plotting functions provide a convenient way to produce simple, high-quality plots, and they can be quite efficient when working with basic **univariate data** involving only one or two variables. However, modern research often demands the handling of far more complex datasets — ones that may include multiple **response variables** alongside numerous explanatory variables. Each additional variable adds layers of complexity and nuance to your data and, by extension, to the plots used to visualize them. While R's built-in plotting tools can accommodate these more elaborate scenarios, doing so often requires a high level of fluency with R's syntax and customization options. For this reason, this book will forgo R's base plotting system in favour of the widely respected *ggplot2* package, a core component of the *tidyverse*. *ggplot2* provides a more consistent and powerful framework for building complex plots, making it the preferred tool for data visualization throughout this text.

The "gg" in *ggplot2* stands for "grammar of graphics," a term borrowed from Leland Wilkinson's influential book of the same name [@Wilkinson2005]. Now, the word "grammar" might dredge up long-buried memories of dull English classes, but fear not. In this context, the term simply emphasizes that *ggplot2* is built on a coherent set of rules for assembling visualizations. This structure allows users to build a wide variety of plots in a consistent, modular way that can be easily tailored to their data and needs. This is a major improvement over many traditional plotting systems, which often require you to awkwardly cram your data into rigid, predefined formats—like trying to fit a square peg (your beautifully weird data) into a round hole (the software's narrow expectations).

The easiest way to understand how *ggplot2* works is to simply dive in and use it. Along the way, we will also learn a little bit more about R and data manipulation. However, a disclaimer is perhaps useful here:


::: callout-warning
This chapter contains a large variety of functions and strategies for plotting data with *ggplot2*. The reader would do well to head the advice provided at the start of @sec-warning.
:::

The first thing to do will be to ensure that *ggplot2* has been installed into our computer's library of packages and loaded so we can access its functions. As mentioned in @sec-tidy_install, if you have installed and loaded the *tidyverse*, this is already done, but if you chose not to do that,[^shame] *ggplot2* can be installed and loaded as a standalone package as well.

[^shame]:Shame on you.

## An example data set: msleep

Before we can plot anything, we need something to plot. In addition to its large set of plotting functions, the *ggplot2* package also provides a few illustrative data sets.[^data_sets] We will work with the `msleep` data set, which provides a variety of measurements relevant to the sleep behaviour of a wide range of mammals. To access the data you need only run the code `msleep` after the tidyverse has been loaded, which will output a $83 \times 11$ data frame.[^tibble_technically] Given the limited space available in the console window, the data frame is going to be truncated substantially. Thus, if you would like to view the entire data set, you can utilize R's `View()` function, which will display the data in a separate spreadsheet style window.

[^data_sets]:Base R comes with a nice collection of data sets as well. To obtain a list you need only run the function `data()`. To obtain the list of data sets for *ggplot2* you need only include the package name as an argument in this function: `data(package = "ggplot2")`

[^tibble_technically]:Technically we are looking at a *tibble*, which is the *tidyverse's* own take on a data frame. For our present purposes though, this is a distinction without a difference.

```r
msleep # print data to console
View(msleep) # view the data in a spreadsheet-style window
```
```{r}
#| echo: false
library(DT)
datatable(msleep, options = list(
  scrollY = "500px",
  scrollX = TRUE,
  paging = FALSE
))
```

Looking closely at the data, we can see a variety of variables (the column names) that are, for the most part, self explanatory. In this case, the column names represent distinct variables that have been measured and, particularly with larger data frames that cannot be adequately printed to the console, it is often useful to have R list out the name of each column. We can do this quite easily using the `names()` function.

```{r}
names(msleep)
```

Now, while the names of each column are self-explanatory, the elements of each column are perhaps less so.  For instance, in the `$sleep_total` column, are we looking at values in minutes, hours, or days? In the `$conservation` column we can see a number of abbreviations such as `lc`, `nt`, `vu`, and so on. What do we make of those? A good starting point for answering these questions is to check the documentation associated with the data set, which all CRAN packages are required to include. This can be accessed in the usual way with a `?`

```r
?msleep
```

Inspecting the documentation, we can see that `$sleep_total` is given in hours and that the column `$conservation` indicates "the conservation status of the animal".  Admittedly, concerning this latter column, that does not tell us too much, but it does at least give us a starting point for understanding what those values might represent.  In all likelihood, we are seeing abbreviations for the IUCN's (International Union for Conservation of Nature) species ranking. 

- lc = Least Concern
- nt = Near Threatened
- vu = Vulnerable
- en = Endangered
- cd = Conservation Dependent

Using a **scatter plot** as a basic starting point, we will graph the relationship between the variables body weight (kg) and sleep total (hours).  These are represented by the columns `$bodywt` and `$sleep_total` respectively. 

## Adding layers

*ggplot2* constructs plots by adding visual layers on top of one another. The first layer is the grid upon which our scatter plot's points will appear.  To generate this first layer we can simply type:

```{r}
ggplot(data = msleep, aes(x = bodywt, y = sleep_total))
```

Looking at the `ggplot()` function we typed, we can see that the argument `data` tells *ggplot2* where the data is coming from - in this case it is coming from the `msleep` data frame.  The `x` and `y` arguments are telling *ggplot2* what variables/columns should be mapped to the $x$ and $y$ axis respectively.  Notice that, not only has *ggplot2* labelled the axis accordingly, but it has also given them scales that correspond to size of the values found in both columns.

Next we will, quite literally, add (`+`) a layer of points on top of this by typing `+ geom_point()`. The term "geom" here is just an abbreviation of "geometric object", and points are one of many different types of geometric object *ggplot2* recognizes.

```{r}
ggplot(data = msleep, aes(x = bodywt, y = sleep_total)) +
  geom_point()
```

At this juncture, it is worth taking a moment to talk about how this code we have written has been organized. Here we placed `geom_point()` on a new line and indented it. This was not something we strictly had to do.  We could have put everything on a single line like so ...

```r
ggplot(data = msleep, aes(x = bodywt, y = sleep_total)) + geom_point()
```

But, particularly as we add more customization to the plot, this style of writing becomes hard to read. The *tidyverse's* <a href = "https://style.tidyverse.org/syntax.html#long-lines" target = "_blank">R style guide</a> recommends that no line of code exceed 80 characters, which is the advice most of the R community adheres to. 

::: {.callout-tip collapse = "true"}
## Displaying the 80-character margin

RStudio and Positron can both be configured to display a vertical margin line at the 80-character limit.

**RStudio:**

1. Select *Tools* from the global menu.
2. Choose *Global Options*.
3. Select *Code* in the sidebar.
4. Choose *Display* and tick the box labelled "Show margin".

**Positron:**

1. Select *File* > *Preferences* > *Settings* from the global menu.
2. In the settings search bar, search for `ruler`.
3. Under *Editor: Rulers*, click the link labelled *Edit in settings.json*.
4. In the file that opens, edit the relevant line to read:
   ```json
   "editor.rulers": [80]
   ```
5. Save the file.
:::

To ensure that you do not exceed limit with larger blocks of code, it is worth remembering that you can always move portions of code to a new line after a comma, operator, or unclosed parentheses. The indentation we used is purely to guide the eye in recognizing that `geom_point()` belongs to a larger block of code.[^indent]


[^indent]:While the R programming language allows users to indent code with reckless abandon, some programming languages, such as *Python*, make it part of their syntax and thus require it to be used in very specific ways.

### Inspecting potential outliers

At present, the plot does not look like much. There are numerous points scattered between 0 and 1000, and a couple of very extreme points beyond which are skewing the $x$-axis scale and making the majority of the data difficult to visualize. Given how rare and extreme these two values appear, we should inspect them to ensure that they are not errors within the data set (i.e., ensure that there is not a 2500 kg mouse, bird, or other such abomination in our data set). To accomplish this, most people will instinctively try to scan the data frame's 83 rows one by one with their eyes. Obviously, that strategy will be slow, inefficient, and highly prone to error. A better strategy is to have R isolate these values using the `filter()` function which is part of the *tidyverse's* *dplyr* package.[^subset_func] We simply give the function our data frame, and then specify a logical rule to subset by. In this case we will tell the function to show us all the rows that have a body weight greater than 2000.

[^subset_func]:Base R has a (more or less) equivalent function `subset()` that we could use as well. There are reasons for preferring `filter()`, but in this context there is no advantage to using either.

```{r}
filter(msleep, bodywt > 2000)
```

A quick glance at the output reveals that these two points represent the Asian and African elephant respectively. Thus, while these values are quite extreme and do not seem to be terribly representative of the data as a whole, they are not mistakes and therefore should remain in the data set.  However, this begs the question, how do we visualize this data adequately with such odd scaling?

### Logarithms

A common strategy in cases like this where larger values tend to become more and more extreme (i.e., exhibit some kind of exponential growth) is to plot the logarithm of the values. As a refresher of high school mathematics, logarithms are essentially exponents in reverse. For example:

$$10^3 = 10 \times 10 \times 10 = 1000$$

A "base-10" logarithm, for instance, simply undoes this process by stating how many $10$s it takes to create $1000$.

$$\log_{10}(1000) = 3$$

A "base-2" logarithm asks: how many $2$s are required to create $1000$?

$$\log_{2}(1000) \approx 9.966$$

Thus, $2^{9.966} \approx 1000$.  

A "natural" logarithm uses a base denoted as $e$ (Euler's Number), which is approximately $2.71828$.[^euler_num]

[^euler_num]:In R you can obtain this by running `exp(1)`.

Base-10, base-2, and natural logarithms represent the most widely used types of logarithms,[^nat_log] but you can technically use any base you desire.  As seen below, the use of logarithms in R is very straightforward.

[^nat_log]:For clarity and consistency the natural logarithm of $1000$ has been written $\log_e(1000)$, but it is common practice to identify natural logarithms using "$\ln$". E.g., $\ln{(1000) \approx 6.908}$.

```{r results = "hold"}
log10(1000) # Base-10 function
log2(1000) # Base-2 function
log(1000) # Natural log
log(1000, base = 666) # Pick your own base
```

A base-10 logarithm is generally considered the most intuitive, so we will use that. There are various ways to incorporate a logarithmic scale on our plot's axis, but perhaps the safest way is to simply add a new column of $\log_{10}$ values to our dataframe and plot that instead of the standard `$bodywt` column (however, see the callout below for a alternative method of scaling the axis).

```{r}
#| source-line-numbers: "1,3-4"
msleep$bodywt_log10 <- log10(msleep$bodywt)

# Re-plot the data
ggplot(msleep, aes(x = bodywt_log10, y = sleep_total)) +
  geom_point()
```

::: {.callout-tip collapse = "false"}
## Transforming the axis, not the data
In the previous example, the logarithm was applied by creating a new column of $x$-axis values and plotting that. However, this means that, if you want to interpret the numbers in their original units, you need to calculate $10^x$, which can be annoying.

An alternative strategy would be to keep the `$bodywt` column as is and just scale the plot's axis itself to increment logarithmically, which *ggplot2* will do straightforwardly.

```{r}
#| source-line-numbers: "3"
ggplot(msleep, aes(x = bodywt, y = sleep_total)) +
  geom_point() +
  scale_x_continuous(trans = "log10")
```

The advantage of this method is you can look at a point's value on the $x$-axis and know immediately that it corresponds to a weight of $x$ kg. The drawback is you may end up with excessively small or large values on the axis, hence the scientific notation you see in the plot.
:::






<br><br><br><br><br><hr><br>
<img src = "imgs/misc/under_construction.gif">

