%Preface
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}
Since prefaces often go unread, I shall keep this brief. This text was born out of a need to offer my students a robust, open-access (i.e., free) introduction to R, specifically for those without any programming experience. Long term it is intended to serve as a practical, open-access manual, guiding complete beginners through both statistics and statistical programming in a thorough and clear manner. Please consider everything herein a work in progress.

\vspace{2em}

{
\headingfont \Large
\noindent Foolish Assumptions Made by Your Author:
\normalsize\normalfont
\begin{itemize}
    \item Students will read this (yes, I'm an optimist).
    \item The reader probably hates math, but realizes that math is like physical exercise - necessary, but rarely enjoyable and often exhausting.
    \item Beginners with R and programming will start this book at the beginning and move sequentially through the topics (because skipping ahead is like summoning a demon without comprehending what you are really doing. You’ll regret it. Of that I am confident.).
    \item Readers will not take the sillier things I say in this book too seriously.
\end{itemize}
}

\vspace{2em}

{
\headingfont \Large
\noindent Errata:
\normalsize\normalfont
\begin{itemize}
\item No manuscript, however cursed or consecrated, is free from the creeping taint of error. Should you unearth any faults, oversights, contradictions, or anomalies—do not consign your discovery to silence. Instead, record your findings within the digital vaults of GitHub, where the keeper of this tome may attempt to contain and exorcise the corruption...before it spreads further.
    \begin{itemize}
        \item \url{https://github.com/statistical-grimoire/book/issues}
    \end{itemize}
\end{itemize}
}


